---
title: "R Notebook"
output: github_document
---

# Preliminary imports

We need to practice our data wrangling and tidying skills, in addition to importing different types of data. Thus, we'll bring in `tidyverse` and `readxl`.

```{r lib_import, message = FALSE }
library(tidyverse)
library(readxl)
library(knitr)
```

# Problem 1

This problem requires us to bring in the NYC Transit dataset and clean up the data. The data is contained within a subdirectory of the project, so we'll pull it from there.

```{r P1_load_data, message = FALSE }
raw_nyc_data = read_csv(file = './data/P1/nyc_subway.csv')
```

`raw_nyc_data` in its current state is far from tidy form. Each route has its own column, and many of the cells in these route columns are missing. Some stations have multiple rows that only differ on the entrance locations, making much of the dataset redundant if we don't care about the entrance details. Lastly, the problem doesn't need many of the columns in the raw dataset, so they can be safely excluded.

As we clean the data, we'll address these issues and end up with a tidy data set.

```{r P1_data_cleaning }
tidy_nyc_data = raw_nyc_data %>% 
  janitor::clean_names() %>% 
  gather(data = ., key = route_number, value = route_served, route1:route11, na.rm = TRUE) %>% 
  mutate(can_enter = ifelse(entry == "YES", TRUE, FALSE),
         has_vending = ifelse(vending == "YES", TRUE, FALSE),
         is_ada = ada) %>% 
  select(., station_name, line, route_served, station_latitude, station_longitude, 
         entrance_type, can_enter, has_vending, is_ada) %>% 
  distinct(., station_name, line, route_served, .keep_all = TRUE) %>%
  arrange(., station_name, line) 
```

## The cleaning process

`tidy_nyc_data` is the result of our tidying process. After fixing up the column names, I gathered all of the route columns so that we get a 1-to-1 correspondence to each station and every route they serve. Gathering the routes removes the multitude of missing values present in the route columns. `entry` and `vending` have been converted into true Boolean variables and have been renamed to be more descriptive of their "yes/no" nature (the same renaming was done to `ada`). I used `select()` to rearrange the columns to give station the main focus, while removing extraneous columns not required in the problem. Since many of the stations have multiple entrances, I used `distinct()` to make sure that each station-to-route pair is unique.

The ending result keeps the station and line information, and each route has its own row. Other station information such as the station location, entrance type, ADA status has been kept. Customers who may need to figure out if they can even enter a particular station or purchase a Metro Card can use the corresponding Boolean columns.

The original dataset had `r nrow(raw_nyc_data)` rows and `r ncol(raw_nyc_data)` columns, while the processed dataset has been reduced to `r nrow(tidy_nyc_data)` rows and `r ncol(tidy_nyc_data)` columns. The data is now in tidy form.

## NYC Transit Data Inquiry

* How many distinct stations are there? Note that stations are identified both by name and by line (e.g. 125th St A/B/C/D; 125st 1; 125st 4/5); the distinct function may be useful here.

If each station is identified by the station and line, then we can use the `distinct()` function again to figure out how many unique station-line values there are.

```{r P1_Q1_answer }
distinct_stations = tidy_nyc_data %>% 
  select(., station_name, line) %>% 
  mutate(station_line = paste(station_name, line)) %>% 
  distinct(., station_line)
```

`distinct_stations` tells us that there are `r nrow(distinct_stations)` distinct stations. However, if we 

* How many stations are ADA compliant?
* What proportion of station entrances / exits without vending allow entrance?
* Reformat data so that route number and route name are distinct variables. How many distinct stations serve the A train? How many are ADA compliant?



## bindrows lets you stack datasets on top of each other